---
title: "In-class template"
author: "An Bui"
format: pdf
---

# 1. Set up

## Packages

```{r}
#| label: packages

# general use and cleaning
library(tidyverse)
library(here)
library(janitor)
library(snakecase)

# wrap axis labels
library(scales)

# reading .xlsx files
library(readxl)

# visualize missing data
library(naniar)

# arrange plots
library(patchwork)

```

## Data

```{r}
#| label: data

# taxonomic information
taxon_list <- read_csv(here("data", "taxon_list.csv"))

# aquatic invertebrates
aq_ins <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Aquatic Insects"
)

# site information
sites <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "sites"
)

# water quality
water_quality <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Water Quality",
  na = c("", "n/a", "N/A", "over", "173+")
)


```

# 2. Cleaning

## `sites` object

Goal: create a new object from `sites` that only includes NCOS sites

```{r}
#| label: ncos-sites

# creating new object from sites
ncos_sites <- sites |> 
  # filtering to only include NCOS quarterly sampling sites
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")
```


## clean `taxon_list`

Goal: convert character strings in the existing `taxon_list` data frame into snakecase

```{r}
#| label: cleaning-taxon

# creating new object from taxon_list
taxon_list_clean <- taxon_list |> 
  # converted taxon names to snake case
  mutate(verbatim_name = to_any_case(verbatim_name, 
                                     case = "snake"))
```

## clean `water_quality`

Goal: create a data frame with median pH, median DO (mg/L), median salinity (ppt) for each date and site

- create a new object from `water_quality`
- clean column names
- only include surveys from quarterly NCOS sites
- create a new column called `date_site` to join with later data frames
- group by date_site
- calculate median pH, salinity, and dissolved oxygen
- ungroup data frame
- filter out any outliers

```{r}
#| label: cleaning-water-quality

# creating a new object from water quality
water_quality_clean <- water_quality |> 
  # cleaning column names
  clean_names() |> 
  # filtering join: only include sites in ncos_sites
  semi_join(ncos_sites, by = "site") |> 
  # creating new column to join with later data frames
  unite("date_site", date, site, remove = TRUE) |> 
  # grouping by date/site
  group_by(date_site) |> 
  # calculating median pH, salinity, and DO
  summarize(med_ph = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  # ungrouping the data frame
  ungroup() |> 
  # filtering out salinity outliers
  filter(date_site != "2022-08-12_NVBR")

```


## clean `aq_ins`

Goal: create a long format data frame in which each row contains the counted individuals in each taxon during each survey. The data frame should include taxonomic information for each organism and the water quality information for each survey

- create a new object from `aq_ins`
- clean column names
- filter to only include quarterly NCOS sites
- select columns
- filter to only include "filtered beaker" sample
- convert the data frame to long format
- group by site, date, and taxon
- calculate average abundance per liter (sum counts divided by 7.5 liters)
- create a new column called `date_site` to join with water quality
- join data frames
- join with taxon list

```{r}
#| label: cleaning-aq-ins

# create a new object
aq_ins_clean <- aq_ins |> 
  # cleaning column names
  clean_names() |> 
  # filtering join: only include quarterly NCOS surveys from ncos_sites
  semi_join(ncos_sites, by = "site") |> 
  # renaming column
  rename(date = date_on_vial) |> 
  # select columns of interest
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae,
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  # filtered to only include filtered beaker samples
  filter(sample_type %in% c("FB250", "FB 250")) |> 
  # make data frame longer
  pivot_longer(
    # columns to make longer
    cols = ostracod:annelida_polychaete,
    # new column for the old column names
    names_to = "taxon",
    # new column for the values in each of the old columns
    values_to = "abundance"
  ) |> 
  # grouped by date, site, and taxon
  group_by(date, site, taxon) |> 
  # calculated average count/liter (total abundance/7.5 L)
  summarize(ave_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  # ungrouped
  ungroup() |> 
  # created new column to join with data frames
  unite("date_site", date, site, remove = FALSE) |> 
  # joined with water quality clean
  left_join(water_quality_clean, by = "date_site") |> 
  # joined with taxonomic information
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  # fill in full site names
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "East Channel",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~ "West Pond",
    site == "NDC" ~ "Devereux Creek"
  )) |> 
  # setting factor levels for full site name by median salinity
  mutate(site_full = fct_reorder(site_full,
                                 med_sal,
                                 .fun = "median",
                                 .na_rm = TRUE),
         site_full = fct_rev(site_full))


```

# 3. Visualizing

## Visualize missing values

```{r}
#| label: missing-viz

gg_miss_var(aq_ins_clean)
```

Missing values for salinity, pH, and dissolved oxygen might influence later figures.

## Visualize salinity

- x-axis: `site_full`
- y-axis: `mean_sal`
- geometry to show salinity during survey: `geom_jitter`
- geometry to show median salinity: `stat_summary`

```{r}
#| label: salinity-plot

# creating a new object to store plot
salinity_plot <- ggplot(data = aq_ins_clean,
                        # x-axis: full site name
                        # y-axis: median salinity
                        mapping = aes(x = site_full,
                                      y = med_sal)) +
  # points representing median salinity at each survey
  geom_jitter(height = 0) +
  stat_summary(geom = "point",
               fun = median,
               color = "red",
               size = 2)

# displaying the object
salinity_plot
```

Sites seem to differ in salinity. North of Venoco Bridge, East Channel, and Main Channel tend to be hypersaline, while North of Phelps Bridge, Phelps Road, and Devereux Creek is the least saline. South of Phelps Bridge tends to be more freshwater but is higher in salinity than the other, lower salinity sites.










